---
title: 'Data Viz: Basic Graphics'
author: "STUDENT NAME HERE"
output: 
  pdf_document:
    toc: yes
    toc_depth: '4'
  html_document:
    collapsed: no
    toc: yes
    toc_depth: 4
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '4'
---

### I. Basic Graphics

We will be doing simplified reproductions of figures found in Storytelling with Data. When we learn more about ggplot we can personalize and polish our plots more.

#### 1) Heatmapped Table

To create a heatmapped table you might want to use the following packages: ztable or tidyverse.

I had a lot of difficult with installing ztable. So I will recreate heatmapped tables using the tiling feature of ggplot2.

##### Load the example data

```{r}
# Create a matrix 
bars<-matrix(c(4, 3, 1, 1,
               5, 6, 3, 1,
               4, 5, 1, 2,
               4, 5, 3, 2,
               7, 6, 5, 3), nrow=5, byrow=TRUE)

# label rows and columns
rownames(bars)<-c("Cat1", "Cat2", "Cat3", "Cat4", "Cat5")
colnames(bars)<-c("A", "B", "C", "D")

bars
```
##### This data is NOT tidy

I use the `melt` function to establish observations are row and variables as columns.  You can also use `gather`.

```{r}
#install.packages("reshape2")
library(reshape2)
barsM<-melt(bars)

# Now look at the data
head(barsM)
```

##### Now create a tiled plot

```{r}
library(tidyverse)
ggplot(barsM, aes(#INSERT CODE HERE#)) + # x and y axes => Var1 and Var2
  #INSERT CODE HERE# + # background colours are mapped according to the value column
  #INSERT CODE HERE# +
  #INSERT CODE HERE#
```
##### Here is another example from the textbook

```{r}
## FIG 2.5 (Data on github)
heat<-read.delim("https://raw.githubusercontent.com/kitadasmalley/DATA502/main/FALL2022/Data/swdFig2.5.txt",
                 header = TRUE)

rownames(heat)<-heat[,1]
heat<-heat[,-1]
headM<-melt(as.matrix(heat))

# Relevel categories in reverse
headM$Var1<-ordered(headM$Var1, levels = rev(levels(headM$Var1)))

ggplot(headM, aes(#INSERT CODE HERE# + # x and y axes => Var1 and Var2
  #INSERT CODE HERE# + # background colours are mapped according to the value column
  #INSERT CODE HERE#+
  #INSERT CODE HERE#+
  #INSERT CODE HERE#
```

```{r}
## NOTE: You can also use the gather function in tidyr to do the manipulation
heat<-read.delim("https://raw.githubusercontent.com/kitadasmalley/DATA502/main/FALL2022/Data/swdFig2.5.txt",
                 header = TRUE)
headG<-heat%>%
  gather("type", "percent", -X)
```


#### 2) Simple Text

Here is a simple example where text would suffice:

```{r}
## RECREATE FIG2.2 (DATA ON GITHUB)
mothers<-read.csv("https://raw.githubusercontent.com/adamribaudo/storytelling-with-data-ggplot/master/data/FIG0202-3.csv",
                header=TRUE, 
                stringsAsFactors = FALSE)

# geom_col used when counts instead of indiv obs
ggplot(mothers, aes(#INSERT CODE HERE#
  #INSERT CODE HERE#
  #INSERT CODE HERE#
  #INSERT CODE HERE#
  #INSERT CODE HERE#
  #INSERT CODE HERE#
```

#### 3) Scatter Plot

##### Import the data

```{r}
## RECREATE FIG2.6 and FIG2.7
## DATA ON GITHUB
drive<-read.csv("https://raw.githubusercontent.com/adamribaudo/storytelling-with-data-ggplot/master/data/FIG0206-7.csv",
               header=TRUE, 
               stringsAsFactors = FALSE)

## USE str to see what data are available
str(drive)
```

##### Remove Currency Symbols

When R sees non-numeric characters anyway in the column it will assume the column is composed of strings.  We use `gsub` as a way to find and replace these characters. 

```{r}
drive$Cost.Per.Mile2<-as.numeric(gsub("[\\$,]", "", drive$Cost.Per.Mile))
drive$Miles.Driven2<-as.numeric(gsub("[\\$,]", "", drive$Miles.Driven))

str(drive)
```

##### Create above average variable

```{r}
above<-drive%>%
  mutate(above=#INSERT CODE HERE#
```

##### Plot your data

```{r}
ggplot(above, aes(#INSERT CODE HERE#
  geom_#INSERT CODE HERE#
  geom_#INSERT CODE HERE#
  geom_#INSERT CODE HERE#
  xlab(#INSERT CODE HERE#
  ylab(#INSERT CODE HERE#
  theme_#INSERT CODE HERE#
  theme(#INSERT CODE HERE#
  ggtitle(#INSERT CODE HERE#
```

#### 4) Line Graph

In this example, we need to use a date type variable. This can be done with the package lubridate.

```{r}
## RECREATE FIG2.8
#install.packages("lubridate")
library(lubridate)

lineEx<-read.csv("https://raw.githubusercontent.com/adamribaudo/storytelling-with-data-ggplot/master/data/FIG0209.csv",
                header=TRUE, 
                stringsAsFactors = FALSE)

lineDate<-lineEx%>% 
  mutate(date = ymd(paste(Year, Month, 1)))

ggplot(lineDate, aes(#INSERT CODE HERE#
  geom_#INSERT CODE HERE#
  theme_#INSERT CODE HERE#
```

#### 5) Slope Graph

Warning: These data are not tidy. Use the gather function!

```{r}
## RECREATE FIG2.10

slope<-read.csv("https://raw.githubusercontent.com/adamribaudo/storytelling-with-data-ggplot/master/data/FIG0210-11.csv",
                header=TRUE, 
                stringsAsFactors = FALSE)

# needs some transformation
colnames(slope)<-c("Item", 2014, 2015)

slopeT<-slope%>%
  gather("year", "percent", -Item)

slopeT$percent2<-as.numeric(gsub("[\\%]", "", slopeT$percent))/100

ggplot(slopeT, aes(#INSERT CODE HERE#
  geom_#INSERT CODE HERE#
  theme_#INSERT CODE HERE#
```

#### 6) Bar Graphs

Let’s create some fake data for this:

```{r}
# This is how the book does it
bars<-matrix(c(4, 3, 1, 1,
               5, 6, 3, 1,
               4, 5, 1, 2,
               4, 5, 3, 2,
               7, 6, 5, 3), nrow=5, byrow=TRUE)

bars<-cbind(c("Cat1", "Cat2", "Cat3", "Cat4", "Cat5"), 
            bars)

colnames(bars)<-c("Category","A", "B", "C", "D")

# But we need to make it tidy
barsT<-as.data.frame(bars)%>%
  gather("Letter", "Count", 2:5)

barsT$Count<-as.numeric(barsT$Count)
```

There are many different kinds of bar graphs: 

##### A. Vanilla (Vertical)

```{r}
ggplot(barsT, aes(#INSERT CODE HERE#
  geom_bar(stat=#INSERT CODE HERE#
```

##### B. Vertical Stacked Bar Chart

```{r}
ggplot(barsT, aes(#INSERT CODE HERE#
  geom_bar(#INSERT CODE HERE#
```

###### B2. Vertical Stacked Bar Chart as Percentages

```{r}
ggplot(barsT, aes(#INSERT CODE HERE#
  geom_bar(#INSERT CODE HERE#
```

##### C. Vertical Side-by-side Bar Chart

```{r}
ggplot(barsT, aes(#INSERT CODE HERE#
  geom_bar(#INSERT CODE HERE#
```

##### D. Horizontal Bar Chart

Any of the plots above can be made horizontal!

```{r}
ggplot(barsT, aes(#INSERT CODE HERE#
  geom_bar(#INSERT CODE HERE#
  coord_#INSERT CODE HERE#
```

### II. Esquisse

Since we are learning how to make graphics for the first time. I suggest using this helpful tool!

```{r}
#install.packages("esquisse")
library(esquisse)
```

#### Please load the data for the examples that I have pre-processed

##### A. Mothers

```{r}
## MOTHER EXAMPLE
mothers<-read.csv("https://raw.githubusercontent.com/adamribaudo/storytelling-with-data-ggplot/master/data/FIG0202-3.csv",
                header=TRUE, 
                stringsAsFactors = FALSE)
```


##### B. Drive

```{r}
## MILES DRIVEN
drive<-read.csv("https://raw.githubusercontent.com/adamribaudo/storytelling-with-data-ggplot/master/data/FIG0206-7.csv",
               header=TRUE, 
               stringsAsFactors = FALSE)
drive$Cost.Per.Mile2<-as.numeric(gsub("[\\$,]", "", drive$Cost.Per.Mile))
drive$Miles.Driven2<-as.numeric(gsub("[\\$,]", "", drive$Miles.Driven))

above<-drive%>%
  mutate(above=(Cost.Per.Mile2>=mean(Cost.Per.Mile2)))
```


##### C. Line Date

```{r}
## PASSPORT WAIT
#install.packages("lubridate")
library(lubridate)

lineEx<-read.csv("https://raw.githubusercontent.com/adamribaudo/storytelling-with-data-ggplot/master/data/FIG0209.csv",
                header=TRUE, 
                stringsAsFactors = FALSE)

lineDate<-lineEx%>% 
  mutate(date = ymd(paste(Year, Month, 1)))
```

##### D. Slope T

```{r}
## EMPLOYEE FEEDBACK
slope<-read.csv("https://raw.githubusercontent.com/adamribaudo/storytelling-with-data-ggplot/master/data/FIG0210-11.csv",
                header=TRUE, 
                stringsAsFactors = FALSE)

# needs some transformation
colnames(slope)<-c("Item", 2014, 2015)

slopeT<-slope%>%
  gather("year", "percent", -Item)

slopeT$percent2<-as.numeric(gsub("[\\%]", "", slopeT$percent))/100
```

### III: Let's Practice

```{r}
### MEALS
meals<-read.delim("https://raw.githubusercontent.com/kitadasmalley/DATA502/main/FALL2022/Data/ch2_meals.txt",
                 header = TRUE)
str(meals)

### HEATMAP
ggplot(meals, aes(#INSERT CODE HERE#
  geom_tile(#INSERT CODE HERE#
  geom_text(aes(#INSERT CODE HERE#
  scale_fill_gradient2(#INSERT CODE HERE#
  theme_#INSERT CODE HERE#

### BAR GRAPH
ggplot(meals, aes(#INSERT CODE HERE#
  geom_bar(#INSERT CODE HERE#
  theme_#INSERT CODE HERE#

### LINE PLOT
ggplot(meals, aes(#INSERT CODE HERE#
  geom_#INSERT CODE HERE#
  theme_#INSERT CODE HERE#
```

