---
title: 'Data Viz: Recreating Shot Plot'
author: "INSERT NAME"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Los Angeles Lakers Shooting Heatmap

In this exercise we will recreate a graphic that describes positions on the basketball court the average points scored from that position and the frequency at which shot are taken from this position. 

### The Data

These data represent the LA Lakers 2009/2010 season and come from http://www.basketballgeek.com.  We are going to access this data from my github. 
```{r}
shots_sum<-read.csv("https://raw.githubusercontent.com/kitadasmalley/DATA502/main/FALL2021/Data/shots_sum.csv",
                header=TRUE)
head(shots_sum)

```

## Step 1: Start with a simple scatterplot
```{r message=FALSE, warning=FALSE}
library(tidyverse)

# Step 1 : Scatterplot
## FILL IN CODE HERE ##
```

## Step 2: Add color
```{r}
# Step 2: Add color
## FILL IN CODE HERE ##
```

## Step 3: Size the bubbles
```{r}
# Step 3: Size bubbles
## FILL IN CODE HERE ##
```

#### What is going on with the big dot in the center? 

Its messing up the sizing!

Let's explore:
```{r}
## FILL IN CODE HERE ##
```

## Step 4: Remove point and re-plot
```{r}
# Step 4: Subset the data to remove the "outlier"
## FILL IN CODE HERE ##

# this can also be done with filer
## FILL IN CODE HERE ##

# graph again
## FILL IN CODE HERE ##
```

## Step 5: Change limits to focus within the 3 point boundary
```{r}
## FILL IN CODE HERE ##
```

## Step 6: X and Y distance should be in same distance
```{r}
# Step 6: x and y distance should be in same distance
## FILL IN CODE HERE ##
```

## Step 7: Change color from green to red
```{r}
# Step 7: Change color from green to red
## FILL IN CODE HERE ##
```

## Step 8: Change size scale
```{r}
# Step 8: Change size scale
## FILL IN CODE HERE ##
```

## Step 9: Polishing
```{r}
# Step 9: Polishing
## FILL IN CODE HERE ##

```

