---
title: 'DATA429/599: Intro to Survey Packages '
output:
  html_document:
    toc: true
    collapsed: false
    toc_float: true
    toc_depth: 5
  pdf_document:
    toc: true
    toc_depth: '5'
---

### Learning Objectives: 

In this lesson students will learn to:

* Use survey packages in R
* Define survey objects 
* Inference for the mean

### The Data

These data were put together from NBA and WNBA records for the 2022-2023 and 2022 seasons, respecitively.

For this activity we are going to pretend that we are doing a survey on professional basketball players.  These data represent only the raw responses to the survey.  These data reflect a subset of the NBA and WNBA 2022 population who respondend to the hypothetical survey.  Note that survey weights have been created to adjust for nonresponse. 

Download the data here: 

```{r}
## DOWNLOAD COMPLETE SURVEY DATA
bballComp<-read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/refs/heads/main/DATA429_599/CODE/bballSampComplete.csv")

str(bballComp)
```

### 1. Survey Packages

Load in survey packages.  

```{r message=FALSE, warning=FALSE}
#### SURVEY PACKAGES
#install.packages("survey")
#install.packages("srvyr")
library(tidyverse)
library(survey)
library(srvyr)
```

The new 'srvyr' package utilizes the 'tidyverse' syntax. 

### 2. Define a Survey Design Object

First, we must define a survey design object by mapping to the column with the final weights.  The fpc argument tells R that the sample is finite and without replacement. This should be one number that denotes how large the sampling frame is. 

```{r}
### SURVEY DESIGN OBJECT

## INSTERT CODE HERE ##
```

### 3. Estimating Means

Weighted graphics

```{r}
### WEIGHTED GRAPHIC
## INSTERT CODE HERE ##

```


```{r}
### MEANS AND SE AND CI

## INSTERT CODE HERE ##

```

Let's confirm our weighted estimate:

```{r}
### CONFIRM
## POINT EST
## INSTERT CODE HERE ##

### COMPARE TO UNWEIGHTED
## INSTERT CODE HERE ##
```

### 4. One Sample T-Test

Test for average field goal percentage being 0.50.

```{r}
## ONE T-TEST
### FGP 
## INSTERT CODE HERE ##

## BACK TRANSFROM
## INSTERT CODE HERE ##
```
### 5. Two Sample T-test to Compare Means

```{r}
## TWO T-TEST
## INSTERT CODE HERE ##
```

