---
title: 'DATA429/599: Stratified Random Sample Estimators'
output:
  html_document:
    toc: true
    collapsed: false
    toc_float: true
    toc_depth: 5
  pdf_document:
    toc: true
    toc_depth: '5'
---

### Learning Objectives

In this lesson students will learn to:

* Construct estimates and confidence intervals for stratified random samples from finite populations


Examples from "Sampling: Design and Analysis" by Sharon Lohr.


### Warm-up / Review

Data from a SRS of 120 golf courses, selected from a list of 14,938 golf courses in the United States, are in the file golfsrs.csv

```{r}
golf<-read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/refs/heads/main/DATA429_599/CODE/golfsrs.csv", 
               header=TRUE)
```


#### Visualize 

Display the data in a histogram for the weekday greens fees for nine holes of golf (variable wkday9).  How would you describe the shape of the data? 

```{r warning=FALSE, message=FALSE}
library(tidyverse)

## HISTOGRAM
## INSERT CODE HERE ##
```


#### Average and SRS Standard Error 

Find the average weekday greens fee to play nine holes of golf and give the SE for the estimate.  


```{r}
## SAMPLE STATS
## INSERT CODE HERE ##

## INSERT CODE HERE ##

## STANDARD ERR
## INSERT CODE HERE ##

## FPC
## INSERT CODE HERE ##

## STANDARD ERROR
## INSERT CODE HERE ##
```

#### Confidence Interval 

Use what you found in Part B to construct a 95\% confidence interval. 

```{r}
## CONFIDENCE INTERVAL
## Z CI
zCritVal<-qnorm(0.975)
## INSERT CODE HERE ##


## NEW T CI
tCritVal<-qt(0.975, df=n-1)
## INSERT CODE HERE ##
```


### EX 1: Publications 

In the previous example from Class 7B the SRS, not all departments were represented.  The following data are from a stratified sample using academic division.  The total number of faculty in Biological Sciences is 102, Physical Science is 310, Social Sciences is 217, and Humanities is 178 in the population.

```{r}
pubStr<-read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/refs/heads/main/DATA429_599/CODE/pubStrat.csv", 
                 header=TRUE)

head(pubStr)
```


#### Stratified Average 

Estimate the average number of referred publications by faculty members in the college, and give the standard error.  Use these to construct a 95\% confidence interval. 

```{r}

### BIO
n1<-sum(pubStr$Biological)
ybar1<-sum(pubStr$Publications*pubStr$Biological)/n1
ybar1
s2_1<-(1/(n1-1))*sum(pubStr$Biological*(pubStr$Publications-ybar1)^2)
s2_1
sqrt(s2_1)

### PHYSICAL
n2<-sum(pubStr$Physical)
ybar2<-sum(pubStr$Publications*pubStr$Physical)/n2
ybar2
s2_2<-(1/(n2-1))*sum(pubStr$Physical*(pubStr$Publications-ybar2)^2)
s2_2
sqrt(s2_2)

### SOCIAL
n3<-sum(pubStr$Social)
ybar3<-sum(pubStr$Publications*pubStr$Social)/n3
ybar3
s2_3<-(1/(n3-1))*sum(pubStr$Social*(pubStr$Publications-ybar3)^2)
s2_3
sqrt(s2_3)

### HUMANITIES
n4<-sum(pubStr$Humanities)
ybar4<-sum(pubStr$Publications*pubStr$Humanities)/n4
ybar4
s2_4<-(1/(n4-1))*sum(pubStr$Humanities*(pubStr$Publications-ybar4)^2)
s2_4
sqrt(s2_4)
```

```{r}
### COMBINE
## INSERT CODE HERE ##

## Y BAR STRAT
## INSERT CODE HERE ##

## VAR Y BAR STRAT
## INSERT CODE HERE ##

## STD ERR
## INSERT CODE HERE ##

## CONF INT
## INSERT CODE HERE ##
```

#### Think about it!

  * How does your result from Part A compare to what you found using the simple random sample? 

  * Did stratification increase precision in this sample?  Explain why you think it did or did not.

```{r}
## DESIGN EFFECT 
## INSERT CODE HERE ##
```

### EX 2: Public Health Research

Probability sampling (random selection) allows researchers to generalize from the sample to the population, and random assignment in experiments allows researchers to infer causation, but how often are these methods employed in public health research?  Hayat and Knapp (2017) drew a stratified random sample of 198 articles from the 547 research articles published in 2013 by three leading public health journals.  For each article, they determined the number of authors, the type of statistical inference used (confidence intervals, hypothesis tests, both, or neither), and whether random selection or random assignment was used.  The data are in file healthjournals.csv.

```{r}
#### PUBLIC HEALTH
ph<-read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/refs/heads/main/DATA429_599/CODE/healthjournals.csv", 
             header=TRUE)

str(ph)
```
 
The population sizes of journals are as follows: 
 
  * American Journal of Public Health (AJPH): 280
  * American Journal of Preventive Medicine (AJPM): 103
  * Preventive Medicine (PM): 164
  
#### Confidence Interval for Proportion

Estimate the percentage of the articles that use random selection.  Calculate the point estimate, standard error, and a 95\% confidence interval. 

##### STEP 1: Summary Data

```{r}
## SUMMARISE
## INSERT CODE HERE ##
```

##### STEP 2: Estimates

```{r}
## P HAT STRAT
## INSERT CODE HERE ##

## SE
## INSERT CODE HERE ##
```

##### STEP 3: Confidence Interval

```{r}
### CONF INT
## INSERT CODE HERE ##

```


#### Confidence Interval for Total

Estimate the total number of articles (from the 547) that use random selection. Calculate the point estimate, standard error, and a 95\% confidence interval. 

##### STEP 1: Total Estimate 

```{r}
### ESTIMATE TOTAL
## INSERT CODE HERE ##
```

##### STEP 2: Variance Estimate 
```{r}
## VAR
## INSERT CODE HERE ##

## SE
## INSERT CODE HERE ##
```

##### STEP 3: Confidence Interval
```{r}
## CONF INT
## INSERT CODE HERE ##

```





