---
title: 'DATA429/599: Intro to Survey Packages '
output:
  pdf_document:
    toc: true
    toc_depth: '5'
  html_document:
    toc: true
    collapsed: false
    toc_float: true
    toc_depth: 5
---

### Learning Objectives: 

In this lesson students will learn to:

* Use survey packages in R
* Define survey objects 
* Inference for the mean

### The Data

These data were put together from NBA and WNBA records for the 2022-2023 and 2022 seasons, respecitively.

For this activity we are going to pretend that we are doing a survey on professional basketball players.  These data represent only the raw responses to the survey.  These data reflect a subset of the NBA and WNBA 2022 population who respondend to the hypothetical survey.  Note that survey weights have been created to adjust for nonresponse. 

Download the data here: 

```{r}
## DOWNLOAD COMPLETE SURVEY DATA
bballComp<-read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/refs/heads/main/DATA429_599/CODE/bballSampComplete.csv")

str(bballComp)
```

### 1. Survey Packages

Load in survey packages.  

```{r message=FALSE, warning=FALSE}
#### SURVEY PACKAGES
#install.packages("survey")
#install.packages("srvyr")
library(tidyverse)
library(survey)
library(srvyr)
```

The new 'srvyr' package utilizes the 'tidyverse' syntax. 

### 2. Define a Survey Design Object

First, we must define a survey design object by mapping to the column with the final weights.  The fpc argument tells R that the sample is finite and without replacement. This should be one number that denotes how large the sampling frame is. 

```{r}
### SURVEY DESIGN OBJECT
bball_des<- bballComp %>%
  as_survey_design(
    weights = wi_star, #WEIGHTS
    fpc = fpcBball # WITHOUT REPLACEMENT
  )

bball_des
```

### 3. Estimating Means

Weighted graphics

```{r}
### WEIGHTED GRAPHIC
ggplot(bballComp, aes(x = FGP, 
                          y = ..density.., 
                          weight = wi_star)) + 
  geom_histogram()

```


```{r}
### MEANS AND SE AND CI

bball_des %>%
  summarize(fgp = survey_mean(FGP,
                              vartype = c("se", "ci")
  ))

```

Let's confirm our weighted estimate:

```{r}
### CONFIRM
## POINT EST
num<-sum(bballComp$wi_star*bballComp$FGP)
den<-sum(bballComp$wi_star)
num/den ### CHECKS OUT 

### COMPARE TO UNWEIGHTED
mean(bballComp$FGP)
```

### 4. One Sample T-Test

Test for average field goal percentage being 0.50.

```{r}
## ONE T-TEST
### FGP 
ttest_ex1 <- bball_des %>%
  svyttest(
    formula = FGP -.50 ~ 0,
    design = .,
    na.rm = TRUE
  )

ttest_ex1

## BACK TRANSFROM
ttest_ex1$estimate +.5
```
### 5. Two Sample T-test to Compare Means

```{r}
## TWO T-TEST
ttest_ex2 <- bball_des %>%
  svyttest(
    formula = FGP ~ League,
    design = .,
    na.rm = TRUE
  )

ttest_ex2
```

