---
title: 'DATA429/599: Survey Analysis for Categorical Responses'
output:
  html_document:
    toc: true
    collapsed: false
    toc_float: true
    toc_depth: 5
  pdf_document:
    toc: true
    toc_depth: '5'
---

### Learning Objectives: 

In this lesson students will learn to:

* Use survey packages in R
* Weight data based on auxiliary information
* Analyze categorical responses 

## Example: Thanksgiving Survey

### 0. The Data: 

We are using the data from fivethirtyeight the Here’s What Your Part of America Eats on Thanksgiving article <https://fivethirtyeight.com/features/heres-what-your-part-of-america-eats-on-thanksgiving/>

The original data set can be found here: <https://raw.githubusercontent.com/fivethirtyeight/data/master/thanksgiving-2015/thanksgiving-2015-poll-data.csv>

The variables for the survey are described here:<https://github.com/fivethirtyeight/data/tree/master/thanksgiving-2015>

I cleaned up the data, selected a subset of variables, and created binary variables. My dataset can be found here: <https://raw.githubusercontent.com/kitadasmalley/FA2020_DataViz/main/data/useThanks.csv>

```{r}
### fivethirtyeight
### THANKSGIVING

useThanks<-read.csv("https://raw.githubusercontent.com/kitadasmalley/FA2020_DataViz/main/data/useThanks.csv",
                    header=TRUE)

names(useThanks)
```



### 1. Weighting the Data

How many people in the sample are from each division? 

```{r}
library(tidyverse)

### SAMPLE DIVISION
## INSERT CODE HERE ##
```

We can use Census data to help re-balance the data. 

```{r}
### CENSUS DATA
popDiv<-data.frame(DivName=c("East North Central",
                             "East South Central", 
                             "Middle Atlantic", 
                             "Mountain", 
                             "New England", 
                             "Pacific", 
                             "South Atlantic", 
                             "West North Central", 
                             "West South Central"), 
                   pop=c(46798649, 
                         18931477,
                         41601787,
                         23811346,
                         14757573,
                         52833604,
                         63991523,
                         21179519,
                         39500457))
```

Create weights: 

```{r}
### WEIGHTS
## INSERT CODE HERE ##
```

Join to the data

```{r}
#### JOIN ON TO THE DATA
## INSERT CODE HERE ##
```


### 2. Survey Object

We will pretend that these data come from a simple random sample:

```{r warning=FALSE, message=FALSE}
library(survey)
library(srvyr)

### MAKE SURVEY OBJECT
### SURVEY DESIGN OBJECT
## INSERT CODE HERE ##
```

Let's check to make sure the weights sum to yield the size of each division

```{r}
### ESTIMATE SIZE
## INSERT CODE HERE ##
```

Look at the national proportions

```{r}
### PROPORTIONS BY DIVISION

## INSERT CODE HERE ##
```

### 3. Investigating Cranberry Sauce

```{r}
### SAMPLE PROPORTIONS
## INSERT CODE HERE ##
```

### 4. Subgroups

Are there differences in this distribution across divisions? 

```{r}
### COMPARING CANNED VS HOMEMADE
## INSERT CODE HERE ##
```

We can visualize these propotions with a filled bar graph

```{r}
## FILLED BAR GRAPH
## INSERT CODE HERE ##
```

We can improve this by ordering the division names

```{}
### REORDER
divCranOrd<-divCran%>%
  filter(cranberry=="Canned")%>%
  arrange(desc(p))


joinThanks$DivName<-factor(joinThanks$DivName, levels=divCranOrd$DivName)

thanks_des<- joinThanks %>%
  as_survey_design(
    weights = weight, #WEIGHTS
    fpc = FPC # WITHOUT REPLACEMENT
  )

thanks_des %>%
  filter(cranberry%in%c("Canned", "Homemade"))%>%
  ggplot(aes(x=DivName, y=weight, fill=cranberry))+
  geom_bar(stat = "identity", position="fill")+
  coord_flip()
```

### 5. Chi-Squared Test

Test: 

```{r}
### CHISQR TEST
## INSERT CODE HERE ##
```


Raw Table: 

```{}
## RAW TABLE
chiCranObs<- thanks_des %>%
  filter(cranberry%in%c("Canned", "Homemade"))%>%
  group_by(DivName, cranberry) %>%
  summarize(Observed = round(survey_mean(vartype = "ci"), 3))

chiCranObs
```

Formatted Table: 

```{}
### FORMATTED TABLE
#install.packages("gt")
library(gt)

chiCranTab<- chiCranObs %>%
  mutate(prop = paste0(
    Observed, " (", Observed_low, ", ",
    Observed_upp, ")"
  )) %>%
  select(DivName, cranberry, prop) %>%
  pivot_wider(
    names_from = DivName,
    values_from = prop
  ) %>%
  gt(rowname_col = "cranberry") %>%
  tab_stubhead(label = "Cranberry")

chiCranTab
```

