---
title: 'DATA429/599: Techniques for Handling Unit Nonresponse'
output:
  pdf_document:
    toc: true
    toc_depth: '5'
  html_document:
    toc: true
    collapsed: false
    toc_float: true
    toc_depth: 5
---

### Learning Objectives: 

In this lesson students will learn to:

* The woes of nonresponse!
* How to use auxiliary data to re-weight survey responses in the face of nonresponse  

### The Data

These data were put together from NBA and WNBA records for the 2022-2023 and 2022 seasons, respecitively.

For this activity we are going to pretend that we are doing a survey on professional basketball players. 

Download the data here: 

```{r}
## DOWNLOAD DATA
bball<-read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/refs/heads/main/ProjectData/bBall2022/bballNR.csv")

str(bball)
```

We will also need the tidyverse to help with some wrangling: 

```{r warning=FALSE, message=FALSE}
library(tidyverse)
```

### The Woes of Nonresponse 

Consider two different oracles that know deterministically if someone is predestined to respond or not respond to a survey. 

#### Oracle 1

```{r}
### LOOK AT THE SUMMARY OF ORACLE 1's DATA
bball%>%
  group_by(ORACLE1)%>%
  summarise(n=n(), 
            avg=mean(Salary))

## IT LOOKS LIKE THERE ISN'T A BIG DIFFERENCE BETWEEN THE GROUPS
```

Since you spoke with this oracle you also know the truth of the universe, the true average salary. 

```{r}
## TRUE VALUE
YU<-mean(bball$Salary)
YU

## VERIFY FROM TABLE ABOVE
(127/635)*6141368+(508/635)*6234332
```


Now, pretend that we take a sample of n=150 from the respondent group many many times.  Doing this we can estimate the amount of bias.  

```{r}
### SIMULATE BIAS
bias<-c()
for(i in 1:1000){
  oracle1_Resp<-bball%>%
    filter(ORACLE1==1)
  thisSamp<-sample(1:dim(oracle1_Resp)[1], 150, replace=FALSE)
  thisYR<-mean(oracle1_Resp$Salary[thisSamp])
  thisBias<-thisYR-YU
  bias<-c(bias, thisBias)
}

bias_df1<-data.frame(run=1:1000, bias)

## APPROX
(127/635)*(6234332-6141368)

ggplot(bias_df1, aes(x=run, y=bias))+
  geom_point()+
  geom_hline(yintercept = 18592.8, color="red")
```

#### Oracle 2

Now, in a parallel universe, we go an talk to a different oracle.  

```{r}
### LOOK AT THE SUMMARY OF ORACLE 2's DATA
bball%>%
  group_by(ORACLE2)%>%
  summarise(n=n(), 
            avg=mean(Salary))

## TRUE VALUE
YU<-mean(bball$Salary)
YU

## IT LOOKS LIKE THERE IS BIG DIFFERENCE BETWEEN THE GROUPS
```

Now again, pretend that we take a sample of n=150 from the respondent group many many times.  Doing this we can estimate the amount of bias.  

```{r}
### SIMULATE BIAS
bias2<-c()
for(i in 1:1000){
  oracle2_Resp<-bball%>%
    filter(ORACLE2==1)
  thisSamp<-sample(1:dim(oracle2_Resp)[1], 150, replace=FALSE)
  thisYR<-mean(oracle2_Resp$Salary[thisSamp])
  thisBias<-thisYR-YU
  bias2<-c(bias2, thisBias)
}

bias_df2<-data.frame(run=1:1000, bias2)

## APPROX
(127/635)*(2059872-22839209)

ggplot(bias_df2, aes(x=run, y=bias2))+
  geom_point()+
  geom_hline(yintercept = -4155867, color="red")

```

#### Moral of the Story

We aren't oracles and we never know the truth.  If the group of nonrespondents is very different from the group of respondents are results will be unreliable and we can make the wrong conclusions!

### Using Auxiliary Data

Sometimes we may know something about the population from which we took the sample.  We can use this to check if our data appears to be representive of these dimensions 

```{r}
#### AUXILLARY DATA
### LEAGUE
LeTab<-table(bball$League)
prop.table(LeTab)

### POSITION
PosTab<-table(bball$SimpPos)
prop.table(PosTab)
```

#### Case 1: SRS in a Perfect World 

In a perfect world, we would have no nonresponse.  The following simple random sample for generated previously in R. 

```{r}
#### SRS - NO NONRESPONSE
## DOES THIS LOOK SAMPLE LOOK REPRESENTATIVE?

ggplot(bball, aes(x=as.factor(SRS_0), fill=League))+
  geom_bar(position="fill")

```

Look at the distribution: 

```{r}
## FILTER
sampSRS_0<-bball%>%
  filter(SRS_0==1)

table(sampSRS_0$League)
prop.table(table(sampSRS_0$League))

## ITS NOT TOO BAD... LET'S KEEP GOING
```


Since we feel satisfied and this comes from an SRS, where all the weights are the same.  We can calculate the sample mean. 

```{r}
### CALCULATE THE MEAN FROM THE RESPONDENTS 

## MEAN
mean(sampSRS_0$Salary) ## ESTIMATE
```

If we are willing to make these assumptions and we have no other information this is the best we can do. 



#### Case 2: SRS with MCAR 

In reality we will have missing data.  First, let's explore data that is Missing Completely At Random!

Now, when a respondent was selected but does not respond they will be represented in the data with an `NA`.

```{r}
#### SRS_MCAR
### LOOK AT COLUMN
head(bball$SRS_MCAR, 20)
```

Let's check the non-response rates across Leagues. 

```{r}
## DOES THIS LOOK SAMPLE LOOK REPRESENTATIVE?
## LET'S LOOK AT THE NONRESPONSE RATES
bball%>%
  filter(is.na(SRS_MCAR)==TRUE | SRS_MCAR==1)%>%
  group_by(League)%>%
  summarise(nonRespR=mean(is.na(SRS_MCAR)))

```

These nonresponse rates are pretty similar.  

```{r}
## DOES THE SAMPLE LOOK REPRESENTATIVE
## REMOVE NR
sampSRS_MCAR<-bball%>%
  filter(is.na(SRS_MCAR)==FALSE)%>%
  filter(SRS_MCAR==1)

table(sampSRS_MCAR$League)
prop.table(table(sampSRS_MCAR$League))
```
If we are willing to move forward, we can calculate the sample mean. 

```{r}
## SAMPLE MEAN
mean(sampSRS_MCAR$Salary) ## ESTIMATE
```

#### Case 3: SRS with MAR

Now it's getting real!  There can be relationships between observed explanatory variables and the propensity to respond.  When this happens, we refer to it as missing at random (MAR).

```{r}
#### SRS_MAR
## DOES THIS LOOK SAMPLE LOOK REPRESENTATIVE?
## LET'S LOOK AT THE NONRESPONSE RATES
bball%>%
  filter(is.na(SRS_MAR)==TRUE | SRS_MAR==1)%>%
  group_by(League)%>%
  summarise(NonRespR=mean(is.na(SRS_MAR)))
```

OH NO!!! It look's like there is differential nonresponse!

```{r}
### OH NO!
## REMOVE NR
sampSRS_MAR<-bball%>%
  filter(is.na(SRS_MAR)==FALSE)%>%
  filter(SRS_MAR==1)

table(sampSRS_MAR$League)
prop.table(table(sampSRS_MAR$League))
```
It would not be wise to simple look at the sample mean here.

```{r}
## ESTIMATE
mean(sampSRS_MAR$Salary)
```

### Weighting Adjustments for Nonresponse

#### A. Weighting Class Adjustments

GOAL: Redistribute weights within classes, such that response propensities vary from class to class but are homogenous within.

STEPS: 

* 1. Divide into weighting classes
* 2. Redistribute sample weights
* 3. Check weight adjustments

##### 0. Prepare Data

Add a column for the original weights. 

```{r}
### WEIGHTING CLASS
### USE LEAGUE AS CLASSES 
selectedMAR<-bball%>%
  select(c("Player", "Salary", "League", "SRS_MAR"))%>%
  filter(is.na(SRS_MAR)==TRUE | SRS_MAR==1)%>%
  mutate(wi=635/150)### ORIGINAL WEIGHT 
  

head(selectedMAR)
```


##### 1. Divide into weighting classes

```{r}
### DIVIDE INTO CLASSES 
weightClasses<-selectedMAR%>%
  group_by(League, SRS_MAR)%>%
  summarise(sumWgt=sum(wi))

weightClasses
```


##### 2. Redistribute Weights

```{r}
### REDISTRIBUTE WEIGHTS
selectedMAR1<-selectedMAR%>%
  filter(SRS_MAR==1)%>%
  mutate(WCA_multiplier=case_when(
    League=="NBA"~ (1+114.3/347.133), 
    League=="WNBA"~ (1+12.7/160.867), 
  ))%>%
  mutate(new_wi=wi*WCA_multiplier)

head(selectedMAR1)
```

##### 3. Check TOTAL

```{r}
### STEP 3: CHECK
sum(selectedMAR1$new_wi)

## CLOSE BUT A TINY BIT OF ROUNDING ERROR
```


##### 4. Weighted Estimate

```{r}
### WEIGHTED ESTIMATE
numerator<-sum(selectedMAR1$new_wi*selectedMAR1$Salary)
numerator
denominator<-sum(selectedMAR1$new_wi)
denominator

### AVERAGE
numerator/denominator ### MUCH BETTER!
```


#### B. Poststratification 

If you know the count within your sample of a characteristic (or even multiple characteristics together) you can create stratification weights. 

##### 1. Distribution in the Sample
```{r}
### POSTSTRATIFIED
sampSRS_MAR%>%
  group_by(League)%>%
  summarise(n=n())

### TRUTH
LeTab
```

##### 2. Poststratification Weights

```{r}
### POST STRAT WEIGHTS
### NBA
467/82
### WNBA
168/38

### ADD POST-STRAT WEIGHTS
#head(sampSRS_MAR)

postStrat_MAR<-sampSRS_MAR%>%
  select(c("Player", "Salary", "League", "SRS_MAR"))%>%
  mutate(wi_star=case_when(
    League=="NBA" ~ 467/82, 
    League=="WNBA" ~ 168/38
  ))

head(postStrat_MAR)
```

##### 3. Don't forget to check your work!

```{r}
## CHECK
sum(postStrat_MAR$wi_star)
```

##### 4. Weighted Estimate

```{r}
### PS WEIGHTED ESTIMATE
numeratorPS<-sum(postStrat_MAR$wi_star*postStrat_MAR$Salary)
numeratorPS
denominatorPS<-sum(postStrat_MAR$wi_star)
denominatorPS

### PS AVERAGE
numeratorPS/denominatorPS ### EVEN BETTER!
```

#### COMPARE

```{r}
### TRUTH
YU

### WEIGHTING CLASS
numerator/denominator

### POSTSTRAT
numeratorPS/denominatorPS
```

