---
title: 'DATA429/599: Intro to Survey Packages '
output:
  html_document:
    toc: true
    collapsed: false
    toc_float: true
    toc_depth: 5
  pdf_document:
    toc: true
    toc_depth: '5'
---

### Learning Objectives: 

In this lesson students will learn to:

* Use survey packages in R
* Define survey objects 
* Inference for the mean

### The Data

These data were put together from NBA and WNBA records for the 2022-2023 and 2022 seasons, respecitively.

For this activity we are going to pretend that we are doing a survey on professional basketball players.  These data represent only the raw responses to the survey.  These data reflect a subset of the NBA and WNBA 2022 population who respondend to the hypothetical survey.  Note that survey weights have been created to adjust for nonresponse. 

Download the data here: 

```{r}
## DOWNLOAD COMPLETE SURVEY DATA
bballComp<-read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/refs/heads/main/DATA429_599/CODE/bballSampComplete.csv")

str(bballComp)
```

### 1. Survey Packages

Load in survey packages.  

```{r message=FALSE, warning=FALSE}
#### SURVEY PACKAGES
#install.packages("survey")
#install.packages("srvyr")
library(tidyverse)
library(survey)
library(srvyr)
```

The new 'srvyr' package utilizes the 'tidyverse' syntax. 

### 2. Define a Survey Design Object

First, we must define a survey design object by mapping to the column with the final weights.  The fpc argument tells R that the sample is finite and without replacement. This should be one number that denotes how large the sampling frame is. 

```{r}
### SURVEY DESIGN OBJECT

## INSTERT CODE HERE ##
```

### 3. Estimating Means

Weighted graphics

```{r}
### WEIGHTED GRAPHIC
## INSTERT CODE HERE ##

```


```{r}
### MEANS AND SE AND CI

## INSTERT CODE HERE ##

```

Let's confirm our weighted estimate:

```{r}
### CONFIRM
## POINT EST
## INSTERT CODE HERE ##

### COMPARE TO UNWEIGHTED
## INSTERT CODE HERE ##
```

### 4. One Sample T-Test

Is the average field goal percent for professional basketball players in the 2022 season $50\%$?  


```{r}
## ONE T-TEST
### FGP 
## INSTERT CODE HERE ##

## BACK TRANSFROM
## INSTERT CODE HERE ##
```
### 5. Two Sample T-test to Compare Means

Is the average field goal percent for professional basketball players in the 2022 season different for NBA and WNBA?  


```{r}
## TWO T-TEST
## INSTERT CODE HERE ##
```


### 6. One Sample T-Test for Proportion

Although the average field goal percentage (FGP) is not $50\%$, there are still some professional basketball players who have an FGP that is greater than $50\%$.  How many are there? 

```{r}
### PROPORTION FGP>50%
## INSTERT CODE HERE ##
```

Suppose that you had a friend who was a big basketball fan and they claimed that $10\%$ percent of players with a FGP greater than $50\%$.  Do we have evidence to support that?

```{r}
### ONE SAMPLE PROP
## INSTERT CODE HERE ##

## BACK TRANSFROM
## INSTERT CODE HERE ##
```



### 7. Two Sample T-Test for Proportions 

A filled bar graph is useful for visualizing differences, while conditioning in the subgroups. 

```{r}
### BAR GRAPH : COMPARE LEAGUES
## WEIGHTED
## INSTERT CODE HERE ##
```

Now let's compare the proportions across the two leagues. 

```{r}
### TABLE
### PROPORTION FGP>50%
### COMPARE NBA and WNBA
## INSTERT CODE HERE ##
```



We can also look at position: 

```{r}
### COMPARE POSITIONS
## INSTERT CODE HERE ##
```

Now make a table:

```{r}
### TABLE
### PROPORTION FGP>50%
### COMPARE SimpPos
## INSTERT CODE HERE ##
```

It looks like there is a difference in point estimates, is this significant? 


### 8. Chi-Squared Test for Independence


We can make a table for this

```{r}
#install.packages("gt")
library(gt)

bball_des<- bballComp %>%
  as_survey_design(
    weights = wi_star, #WEIGHTS
    fpc = fpcBball # WITHOUT REPLACEMENT
  )

### TABLE
chi_CI <- bball_des %>%
  mutate(FGP50=(FGP>.50))%>%
  group_by(SimpPos, FGP50) %>%
  summarize(Observed = round(survey_mean(vartype = "ci"), 3))

chi_CI 

### FORMAT INTO TABLE

chi_CI_tab <- chi_CI  %>%
  mutate(prop = paste0(
    Observed, " (", Observed_low, ", ",
    Observed_upp, ")"
  )) %>%
  select(FGP50, SimpPos, prop) %>%
  pivot_wider(
    names_from = SimpPos,
    values_from = prop
  ) %>%
  gt(rowname_col = "SimpPos")%>%
  tab_stubhead(label = "SimpPos")

chi_CI_tab
```

We could also do a chi-squared test of homogeneity: 

```{r}
### CHI SQR
### LEAGUE vs FGP50
## INSTERT CODE HERE ##
```

