---
title: 'DATA429/599: Item Nonresponse and Imputation '
output:
  html_document:
    toc: true
    collapsed: false
    toc_float: true
    toc_depth: 5
  pdf_document:
    toc: true
    toc_depth: '5'
---

### Learning Objectives: 

In this lesson students will learn to:

* Visualize missing data
* Implement imputation methods

### The Data

These data were put together from NBA and WNBA records for the 2022-2023 and 2022 seasons, respecitively.

For this activity we are going to pretend that we are doing a survey on professional basketball players.  These data represent only the raw responses to the survey.  These data reflect a subset of the NBA and WNBA 2022 population and also reflects nonresponse. 

Download the data here: 

```{r}
## DOWNLOAD RAW SURVEY DATA
bballSurv<-read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/refs/heads/main/DATA429_599/CODE/raw_bballSurvey.csv")

str(bballSurv)
```

### 1. Exploring the Data

Let's to a quick EDA to look for patterns.  How about a pairs plot?!

```{r message=FALSE, warning=FALSE}
library(tidyverse)

### PAIRS PLOT
#install.packages("GGally")
library(GGally)

ggpairs(bballSurv, columns = c(5, 7:18))
```

The leagues have different styles of game play which might impact the relationships between the variables. 

```{r, message=FALSE, warning=FALSE}
### ADD COLOR
### INSERT CODE HERE ###
```

#### What happens if we wanted to model salary?

Consider the following two models. 

##### Model 1: 

```{r}
#### WHAT HAPPENS WITH MISSING DATA
### SALARY AS A FUNCTION OF PTS AND LEAGUE
mod1<-lm(Salary~PTS+League, data=bballSurv)
summary(mod1)
```

##### Model 2: 

```{r}
### SALARY AS A FUNCTION OF PTS AND LEAGUE AND AST
mod2<-lm(Salary~PTS+AST+League, data=bballSurv)
summary(mod2)
```


These models are not actually comparable due to missing data!  If any row in the data has a missing value for variables used in the model, that whole line will be removed.  This is called listwise deletion.  

### 2. Visualizing Missing Data

#### A. Using 'dplyr' 

In our first attempt, we might be curious about how pervasive nonresponse is. 

First, let's try to understand how complete the records are.  How many nonresponses does each individual have? 

```{r}
### NONRESP BY PLAYER
nonR<-bballSurv%>%
  is.na()%>%
  rowSums()

### MAKE A NEW DATAFRAME
playerNR<-bballSurv%>%
  select("Player")%>%
  cbind(nonR)%>%
  mutate(respR=(14-nonR)/14)

### PLOT THE FREQ
### COMPLETE >= 80%
### PARTIAL [50, 80)
### BREAKOFF < 50%
ggplot(data=playerNR, aes(x=respR))+
  geom_histogram()+
  geom_vline(xintercept = .8, color="blue", lwd=1.5, lty=2)+
  geom_vline(xintercept = .5, color="red", lwd=1.5, lty=2)+
  theme_bw()
```

Now, lets look at the frequency of nonresponse across the variables. 

```{r}
### NONRESP BY VAR
## INSERT CODE HERE ##
```
Do any variables seem to get missed significantly more than others? 

#### B. Visualize Missing Data

Let's make a graphic to see where the missing data are. 

```{r}
### VISUALIZE MISSING
#install.packages("visdat")
library(visdat)

## INSERT CODE HERE ##
```
#### C.  Patterns in Missing Data

Sometimes there are relationships between variables such that there is a higher occurrence of multiple variables missing at the same time. 

```{r}
### VIM
#install.packages("VIM")
library(VIM)

## INSERT CODE HERE ##
```

It's hard to see with this many variables, let's just look at a select few:

```{r}
### MAYBE WE SHOULD LOOK AT LESS VARIABLES
## SALARY, PTS, GP, GS, MP, FG, AST 

## INSERT CODE HERE ##

```

### 3. Imputation 

Imputation is the practice of using statistical methods to fill in missing data. 

#### A. Mean Imputation

Mean imputation replaces missing values with the average.  

```{r}
#### MEAN IMPUTATION
## CREATE INDICATORS
## REPLACE WITH MEANS

## INSERT CODE HERE ##

### MARGIN PLOT
## INSERT CODE HERE ##
  
## THIS IS NOT GREAT
## THERE IS NOT ENOUGH VARIABILITY 
```
#### B. HOT DECK

Hotdeck imputation takes "donor" from previous complete data records. 

```{r}
#### VANILLA HOT DECK
## INSERT CODE HERE ##

### MARGIN PLOT
## INSERT CODE HERE ##
```

The problem here is that this just borrows the previous value, but often due to relationships/correlations between variables, we should reorder the data to borrow a "better" value. 

```{r}
## PUT IN ORDER (NUM) - GP
## INSERT CODE HERE ##

## MARGIN PLOT
## INSERT CODE HERE ##
```

Similarly, there can be similar values within groups.  

```{r}
## PUT IN GROUPS (SimpPos)
## INSERT CODE HERE ##

## MARGIN PLOT
## INSERT CODE HERE ##
```

#### C. K-Nearest Neighbors

Consider (completed) respondents similar to the one with missing data.  Then average their responses. 

```{r}
### KNN (5)
## INSERT CODE HERE ##

### MARGIN
## INSERT CODE HERE ##
```

We could also do a weighted average so that the neighbors don't have equal weight.

```{r}
## KNN - WEIGHTED

## INSERT CODE HERE ##

### MARGINAL
## INSERT CODE HERE ##
```

