---
title: "DataViz: Alternative to Pie Charts"
author: "INSERT NAME HERE"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Commuting Data

The following data show commuting transportation methods for states on the West Coast of the United States.

```{r warning=FALSE, message=FALSE}
library(tidyverse)

#rds is an special r dataframe
commute <- readRDS("~/Desktop/DataViz Reading/Week 5 Data and Code/commute.rds")

# a subset of states for class
states <- c("or", "ak", "ca", "wa", "id", "nv")
commute_nw <- filter(commute, state %in% states)

```

## Pie Chart

Pie charts are common, but not great for: 

  1) Many categories
  2) Comparison

```{r}
ggplot(commute_nw) +
  geom_bar(aes(x = "", y = prop, fill = factor(transport_type)), 
    stat = "identity", width = 1) +
  facet_wrap(~ state_name) +
  coord_polar(theta = "y") +
  theme_minimal(18) + 
  xlab("") + ylab("") +
  theme(axis.text = element_blank())+
  scale_fill_discrete(name = "Transporation")+
  labs(title="People in Oregon commute by bicycle more \nthan in other (nearby) states")
```

## Activity

### Step 1: Step 1: Sketch an alternative display of the data that would be an improvement.

  * Graphs should employ perceptual tasks as high on the ranking as possible.
  * Keep important comparisons close.

### Step 2:  Share your sketch with your group

  * Discuss what you think was most effective and what was not
  * Choose one idea to implement together in `ggplot2`

```{r}
## INSERT YOUR CODE HERE
```

### Step 3:  Share with class and submit
