---
title: 'DATA151: Diamonds Drill'
author: "STUDENT NAME"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Diamonds Are Forever

### Load the package

```{r warning=FALSE, message=FALSE}
library(tidyverse)
```

#### Learn about the data

```{r}
data("diamonds")

## INSERT YOUR CODE HERE ##
```


### Question 1

Make a new data set that has the average depth and price of the diamonds in the data set. 

```{r}
## INSERT YOUR CODE HERE ##
```

### Question 2

Add a new column to the data set that records each diamond's price per carat.

```{r}
## INSERT YOUR CODE HERE ##
```

### Question 3

Create a new data set that groups diamonds by their cut and displays the average price of each group. 

```{r}
## INSERT YOUR CODE HERE ##
```

### Question 4

Create a new data set that groups diamonds by color and displays the average depth and average table for each group.

```{r}
## INSERT YOUR CODE HERE ##
```

### Question 5

Which color diamonds seem to be largest on average (in terms of carats)?

```{r}
## INSERT YOUR CODE HERE ##
```

### Question 6

What color of diamonds occurs the most frequently among diamonds with ideal cuts?

```{r}
## INSERT YOUR CODE HERE ##
```

### Question 7

Which clarity of diamonds has the largest average table per carats?

```{r}
## INSERT YOUR CODE HERE ##
```

### Question 8

What is the average price per carat of diamonds that cost more than $10,000?

```{r}
## INSERT YOUR CODE HERE ##
```

### Question 9

Of the diamonds that cost more than $10,000 what is the most common clarity?

```{r}
## INSERT YOUR CODE HERE ##
```

### BONUS CHALLENGE

Add two columns to the diamonds data set. The first column should display the average depth of diamonds in the diamond's color group. The second column should display the average table of diamonds in the diamonds color group.

Hint: Use left_join()

```{r}
## INSERT YOUR CODE HERE ##
```
