---
title: 'FALL 2023 DATA151: Midterm #2 (R Section)'
author: "INSERT STUDENT NAME HERE"
date: "TODAYS DATE"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# I Eat DATA for Breakfast!

Please read ALL descriptions and directions. 

### Description: 

A nutritionist at the Food and Drug Administration is studying the effects of cereal marketing on family meal
choices. In particular, she would like to understand how cereal manufacturers market their products in
grocery stores. She became interested in doing this study after noticing how the
cereal was being restocked one day in her local grocery store. The store personnel
were restocking the cereal shelves based on a reference sheet that told them
where everything was to be placed. The placement of each cereal brand seemed
very deliberate. 

To gather data for her study, the nutritionist goes to the local grocery store and records data about cereal nutritional claims and shelf location for 77 cereals. 


#### Citation: 

* Data from Kaggle: https://www.kaggle.com/datasets/crawford/80-cereals

* Motivation modified from: https://www.hawkeslearning.com/Statistics/chapter_project_pdfs/DIS3/ChapterProject_TheBreakfastofChampions_Student.pdf

## Getting Started (5 points - 1 point each!)

* Create a new R markdown document for the Midterm template 
* Create a folder called "MIDTERM_2"
* Save new Rmd in your R Studio Cloud as "data151_mid2_YOURNAME"
* Change the author and date in the heading of the R markdown

## Part 0: Call in the Data and Tidyverse (5 points)

* (Provided) Import the data from Prof. Smalley's github 
* Look at the structure of the data

```{r}
## IMPORT DATA
cereal<-read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/main/DATA151/FA2023/Professor/DATA/cereal.csv",
                 header=TRUE)

## CHANGE SHELF LOCATION TO FACTOR
cereal$shelf<-as.factor(cereal$shelf)
```

#### Look at the structure of your data. (3 points) 

```{r}
### INSERT YOUR CODE HERE ###
```

#### Call in the `tidyverse` (2 points)

```{r}
### INSERT YOUR CODE HERE ###
```


## Part 1: Cereal Brands (5 points and Extra Credit)

Seven cereal brands (manufacturers) are represented in these data: 

* A = American Home Food Products;
* G = General Mills
* K = Kelloggs
* N = Nabisco
* P = Post
* Q = Quaker Oats
* R = Ralston Purina

#### EXTRA CREDIT: 3 points

It would be more useful to have labels within the data frame.  Use the data frame created by Prof Smalley and join it onto the main `cereal` data frame.  After you have joined on the labels you can continue to work with the resulting dataset for the remainder of the midterm.   

NOTE this is not needed to complete the rest of the midterm; however, it does make it nicer to work with.  

```{r}
mfrLabels<-data.frame(mfr=c("N", "Q", "K", "R", 
                            "G", "P", "A"), 
                      manufacturer=c("Nabisco", "Quaker Oats", 
                                     "Kelloggs", "Ralston Purina", 
                                     "General Mills", "Post", 
                                     "American Home Food Products"))

### INSERT YOUR CODE HERE ### 

```



#### Question 1: 5 points

Count how many cereals there are for each brand and order them by most to least.  Which brand produces the most cereals? 

```{r}
### INSERT YOUR CODE HERE ### 
```

## Part 2: Daily Potassium (5 points)

Based on a 2000 calorie diet, the recommended Daily Value (DV) of Potassium is 4,700 milligrams (mg) per day.  

#### Question 2: 5 points

Create a new data frame that has a new column that represents the percent of the Daily Value of Potassium in one serving of the cereal.  The amount of potassium in the data set is in milligrams.  

Which cereal has the highest daily value percent of potassium per serving?  

```{r}
### INSERT YOUR CODE HERE ###
```

## Part 3: Total Carbohydrates (5 points)

Total carbohydrate on the label includes all three types of carbohydrate: sugar, starch and fiber.

#### Question 3: 3 points 

Create a new data frame that contains a new variable, call it `totalCarbs`, which is the sum of `fiber`, `carbo`, and `sugars`.  

Which cereal has the least total carbohydrates per serving? 

```{r}
### INSERT CODE HERE ###
```

#### Question 4: 2 points

Use the data frame you create in the previous question, add a column that represents the ratio of fiber to total carbohydrates.  

```{r}
### INSERT CODE HERE ###
```

## Part 4: Distribution of Rating (5 points)

#### Question 5: 5 points

Describe the shape of the distribution for the consumer ratings (`rating`).  Use the most appropriate geometries to create plot(s).

Be sure to include commentary on symmetry/skew, modality, spread, and outliers. 

```{r}
### YOUR ANSWER HERE ###

```

## Part 5: Shelf Location (5 points)

It is commonly thought that cereals are displayed in certain shelf locations at a market to draw the attention of children.  In this dataset, the variable `shelf` is the display shelf (1, 2, or 3, counting from the floor).  

#### Question 6: 2 points

Make a hypothesis about shelf location (shelf) and the sugar content in a serving of cereal.

```{}
YOUR ANSWER HERE
```

#### Question 7: 3 points

Create a side-by-side boxplot to show the distribution of sugar content across shelf location.  Use the proper aesthetic so that each box contains the color for the shelf location. 

State your observations. 

```{r}
### INSERT YOUR CODE HERE ###
```

