---
title: "Fall 2024 - Midterm #1"
author: "INSERT NAME HERE"
date: "2024-10-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 3: Recreate


#### Step 1: Read the article 


“Mistakes, we’ve drawn a few” - You can access this article on CANVAS.

This is a short article about visualizations published by The Economist and how they have critiqued their own graphics.  I think this is a great lesson about taking a critical lens on our own work and that this is a learning process! 

#### Step 2: Write a 3-5 sentence summary of your personal take-aways from the article (3 points)

Yes, these can be bullet points!

#### Step 3: Choose a graphic to recreate! 

There are two options!  Within each of these options you can either choose to recreate the original graphic (on the left) or the improved graphic (on the right). 

#### OPTION #1 DATA

```{r}
pensionHKS<-read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/refs/heads/main/DATA502/FALL2024/Week7/pensionGapminder.csv", 
                   header=TRUE)

str(pensionHKS)
```


#### OPTION #2 DATA

```{r}
balanceHKS<-read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/refs/heads/main/DATA502/FALL2024/Week7/balances.csv", 
              header=TRUE)

str(balanceHKS)


econPal1<-c("#8ea0ac", "#b4bc41", "#f15b3d", 
            "#008a84", "#70babb", "#eea41d", 
            "#6ecef4", "#e99273", "#00a3da", "#00506d")
```
