---
title: 'DATA152: Motivating Examples'
author: "STUDENT NAME"
output:
  pdf_document:
    toc: yes
    toc_depth: '5'
---

### Learning Objectives

Students will learn how to...

* Create a tidy data frame in R 
* Use the tidyverse to create exploratory graphics 
* Evaluate arguments based on numerical evidence



### Example 1: Is Yawning Contageous?

MythBusters was a show on Discovery Channel from 2003 to 2016.

Each episode typically focused on investigating “popular beliefs, internet rumors, or other myths” by designing experiments and collecting data. 

In Season 3 Episode 4 (aired March 9, 2005), the MythBuster team explored “Is Yawning Contagious” 


Data Source Citations: 

* Information about episode: <https://www.imdb.com/title/tt0768479/>


#### Step 0: Tidyverse

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

#### Step 1: Load the data into R. Create a data frame.

```{r}
### MYTHBUSTERS

### TREATMENT COLUMN
trt<-c(rep("Control", 12), 
       rep("Control", 4), 
       rep("Seeded", 24), 
       rep("Seeded", 10))

### YAWN (RESPONSE) COLUMN
## INSERT CODE HERE ##

### MAKE A DATAFRAME
## INSERT CODE HERE ##

### LOOK AT THE STRUCTURE
## INSERT CODE HERE ##
```

#### Step 2: Look at the data! 

Let's look at bar graphs, which one best tells the story of these data.

```{r}
## BAR GRAPH
## INSERT CODE HERE ##

## STACKED BAR
## INSERT CODE HERE ##

## FILLED BAR
## INSERT CODE HERE ##
```

#### Step 3: Evaluating Numerical Evidence

* 1. What percent of participants in the treatment group yawned?

* 2. What percent experienced yawned in the control group?

* 3. In which group did a higher percentage of participants yawn?

* 4. What is one other possible explanation for the observed difference between the percentages?

### Example 2: Diabetes 

These data represent diabetes test results and biometrics on women of Pima Indian (Akimel O'odham) heritage, who were at least 21 years of age.

Data Source Citations: 

These data come from the pdp R package. 

#### Step 0: Installing the Package and Calling the Library

```{r}
#install.packages("pdp")
library(pdp)
```

#### Step 1: Calling the Data

```{r}
## DATA IS CALLED 'pima'
## INSERT CODE HERE ##
```

#### Step 2: Explore the Data

Boxplot: 

```{r}
## BOXPLOT
## INSERT CODE HERE ##
```

Density: 

```{r}
## DENSITY
## INSERT CODE HERE ##
```

Scatter plot: 

```{r}
## SCATTERPLOT
## INSERT CODE HERE ##
```

