---
title: 'DATA151: Tables and Bars (Part I)'
author: "STUDENT NAME HERE"
output:
  html_document:
    collapsed: no
    toc: yes
    toc_depth: 5
    toc_float: yes
---

### Learning Objectives

In this lesson students will learn to apply categorical data analysis methods to data sets with fundamentally different structures. 

* Work with cross-tabulated data
* Create univarite tables to show marginal distributions 
* Create two-way tables to show joint and conditional distributions
* Create bar graphs and assess which type of bar graph is best for a given scenario (stacked, dodged, filled)

The `tidyverse` package is needed for these examples

```{r warning=FALSE, message=FALSE}
library(tidyverse)
```


### Example #1: Titanic Data

#### Step 0: Install the package

```{r}
#install.packages("titanic")
library(titanic)
```

#### Step 1: Load the data

```{r}
data("Titanic")

str(Titanic)
```

#### Step 2: Reformat as data.frame

If we want to be able to use all our helpful `dplyr` verbs, we need to reformat this table as a `data.frame`.

```{r}
Titanic<-as.data.frame(Titanic)
```

#### Step 3: One-way Table

Create a one-way frequency table for the distribution of class.

```{r}
## frequency table for class
## can you think of how you would do this in dplyr?
titanClass<-Titanic%>%
## YOUR CODE HERE ##

titanClass
```

##### Relative Frequency Table

We might also want to display proportions.

```{r}
## What could we do if we want proportions?

titanClassProp<-titanClass%>%
## YOUR CODE HERE ##

titanClassProp
```

#### Step 4: Univariate Bar Graphs

Let's visualize this distribution. 

##### A. Simple 

Since the data are already cross-tabulated we must specify the height of the bars with `y=Freq` and `stat = "identity"`.

```{r}
## bar graphs
## YOUR CODE HERE ##
```


##### B. Color

Since bars are two dimensional the `color` aesthetic only outlines bars. 

What is going on in this graph?

```{r}
## add color
## YOUR CODE HERE ##
```


##### C. Fill

```{r}
## oops! let's use fill
## YOUR CODE HERE ##
```

##### D. Proportions

If we want the height of the bar to be a non-integer value (such as proportions) we can use `geom_col`.

```{r}
## change y-axis
## YOUR CODE HERE ##
```

##### E. Recipe for a Pie Graph

###### Step 1. 

Make a stacked bar graph.

```{r}
## Pie chart
## 1) Start with a stacked bar
## YOUR CODE HERE ##
```

###### Step 2. 

Use polar coordinates

```{r}
## 2) plot it in a circle
ggplot(Titanic, aes(x=1, y=Freq, fill=Class))+
  geom_bar(stat = "identity")+
  coord_polar("y", start=0)+
  theme_void()
```

#### Step 4: Two-way Table

```{r}
## Joint distributions
## YOUR CODE HERE ##
```

#### Step 5: Types of Distributions

##### A. Joint Distribution

```{r}
## Now Joint
titanJoint<-titan2Way%>%
## YOUR CODE HERE ##

sum(titanJoint$joint)
```

##### B. Marginal Distribution

```{r}
## Marginal Class
titanMargClass<-titanJoint%>%
## YOUR CODE HERE ##
titanMargClass

## Marginal Survival
titanMargSurv<-titanJoint%>%
## YOUR CODE HERE ##
titanMargSurv
```

##### C. Conditional Distribution

```{r}
## Conditional on Class
titanCondClass<-titan2Way%>%
## YOUR CODE HERE ##

titanCondClass
```


#### Step 6: Bar Graphs with Two Variables

##### A. Stacked (default)

```{r}
## stacked
## YOUR CODE HERE ##
```

##### B. Side-by-side (dodge)

```{r}
## side-by-side
## YOUR CODE HERE ##
```

##### C.  Filled

```{r}
## filled
## YOUR CODE HERE ##
```

##### CAUTION: Pies

It is cognitively very difficult to compare pies!

##### Comparing Across Survival 

```{r}
## comparing pies
ggplot(Titanic, aes(x=1, y=Freq, fill=Class))+
  geom_bar(stat = "identity", position = "fill")+
  facet_grid(.~Survived)+
  coord_polar("y", start=0)+
  theme_void()
```



##### Comparing Across Class

```{r}
## USING FACET_GRID
ggplot(Titanic, aes(x=1, y=Freq, fill=Survived))+
  geom_bar(stat = "identity", position = "fill")+
  facet_grid(.~Class)+
  coord_polar("y", start=0)+
  theme_void()

## USING FACET_WRAP
ggplot(Titanic, aes(x=1, y=Freq, fill=Survived))+
  geom_bar(stat = "identity", position = "fill")+
  facet_wrap(.~Class, nrow=2, ncol=2)+
  coord_polar("y", start=0)+
  theme_void()
```

### Activity: FiveThiryEight

### A. Read the Article 

*"Voter Registrations Are Way, Way Down During the Pandemic"* (Jun 26, 2020) by Kaleigh Rogers and Nathaniel Rakich

https://fivethirtyeight.com/features/voter-registrations-are-way-way-down-during-the-pandemic/ 


### B. Discuss in Small Groups 

1. How are graphics used to tell the author’s story?

2. What geometries are used?

### C. The Data

What does the raw data look like?

```{r}
# Import data
vreg<-read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/voter-registration/new-voter-registrations.csv",
               header=TRUE)


head(vreg)

```

### D. Processing the Data

#### Relevel 

Relevel the data so that its in the right order: 

```{r}
# Level the Month variable so that its in the right order (ie not alphabetical)
vreg$Month<-factor(vreg$Month,
                   levels=c("Jan", "Feb", "Mar", "Apr", "May"))

```


#### Tidy

```{r}
### USE spread() FROM tidyr
vregYear<-vreg%>%
  spread(Year, New.registered.voters)

### RENAME THE COLUMNS
colnames(vregYear)<-c("Jurisdiction", "Month", "Y2016", "Y2020")


```


#### Mutate

Add change in registration. 

```{r}
### mutate() FROM dplyr()
vregChange<-vregYear%>%
  mutate(change=Y2020-Y2016)

```


### E. Recreate the graphic

<https://fivethirtyeight.com/wp-content/uploads/2020/06/rogers-rakich.VOTER-REGISTRATION.0622.png?w=575>


```{r}
# type code/answer here

```

#### Other hints
* You can add another column to define color
* Pay careful attention to the axes, you might want to read the help file for facet_wrap


