---
title: 'Introduction to ggplot2: Points'
author: "INSERT NAME"
output: html_document
---


In this lab we will be learning about some basics with `ggplot2`. 

## Part I: Learning about your data

### Step 1: Load in the data

First let's load in the `diamonds` dataset.  This data set is in the `tidyverse` package, so make sure that that library is called first. 

```{r warning=FALSE, message=FALSE}
library(tidyverse)
data("diamonds")
```

### Step 2: Learn a little bit about this data 

The `str` function allows you to learn about the structure of a dataset.  

```{r}
str(diamonds)
```

### It's Your Turn! Learning by doing

1. How many rows are in `diamonds`?  How many columns?  
2. Which variables in `diamonds` are categorical? Which variables are continuous?
(Hint: look at the output for `str()`)
3. What does the `table` variable describe?  Read the help for `?diamonds` to find out. 

## Part II: Start with a basic scatterplot

Here is a simple scatterplot of `price` vs `carat`:

```{r warning=FALSE, message=FALSE}
ggplot(data=diamonds, aes(x=carat, y=price))+
  geom_point()
```

What do you observe? 

### It's Your Turn! Learning by doing

1. Run `ggplot(data=diamonds)`.  What do you see?
2. Make a scatterplot of `price` vs `depth`.
3. What happens if you make a scatterplot of `cut` vs `clarity`? Why is the plot not useful?


## Part III: Aesthetic Mappings

Aesthetic mappings translate 

### A) Color

#### Color Gradient for Ordinal Data
```{r}
# If using a categorical variable each category will have a color
## INSERT YOUR CODE HERE ##
```

#### Unique colors for Nominal Data 
```{r}
# if not ordered..
## INSERT YOUR CODE HERE ##
```

#### Saturation Gradient for Numeric
```{r}
# If using a numeric variable there will be a color gradient 
## INSERT YOUR CODE HERE ##
```

You can also apply a single color to all the data points by specifying the `color` outside of the aesthetic mapping. 

```{r}
## INSERT YOUR CODE HERE ##
```

### B) Transparency

```{r}
## INSERT YOUR CODE HERE ##
```


### C) Shape

```{r}
## INSERT YOUR CODE HERE ##
```


### D) Size

```{r}
## INSERT YOUR CODE HERE ##
```


### It's Your Turn! Learning by doing

1. What’s gone wrong with this code? Why are the points not blue?

```{r}
ggplot(diamonds, aes(carat, price, color="blue"))+
  geom_point()
```

2. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

3. What happens if you map the same variable to multiple aesthetics?

4. What happens if you map an aesthetic to something other than a variable name, like `aes(colour = carat < 3)`? Note, you’ll also need to specify x and y.

## Part IV: Facets

Sometimes it's useful to look at subgroups within our data.  We can do this with facets. 

### `facet_wrap()`

You can specify a single discrete variable to facet by and R organize plots to fill the space. 

```{r}
## INSERT YOUR CODE HERE ##
```

### `facet_grid()`

You can also create a grid of graphs.  The first argument to the function specifies rows and the second columns. 

```{r}
## Grid
## INSERT YOUR CODE HERE ##
```

If you prefer to not facet in the rows or columns dimension, use a . instead of a variable name, e.g. `+ facet_grid(. ~ color)`.

### It's Your Turn! Learning by doing

1. What happens if you facet on a continuous variable?

2. What plots does the following code make? What does `.` do?

```{r eval=FALSE}
ggplot(diamonds, aes(carat, price))+
  geom_point()+
  facet_grid(color~.)

ggplot(diamonds, aes(carat, price))+
  geom_point()+
  facet_grid(.~cut)
```

3. When using `facet_grid()` you should usually put the variable with more unique levels in the columns. Why?