---
title: '(DATA151) EDA with Palmer Penguins'
output:
  html_document:
    toc: true
    collapsed: false
    toc_float: true
    toc_depth: 5
  pdf_document:
    toc: true
    toc_depth: '5'
---

### Learning Objectives

In this lesson students will learn to apply categorical data analysis methods to data sets with fundamentally different structures. 

* Wrangle different types of data using base R and Tidyverse functions in R.
* Implement exploratory data analysis techniques in R, including the following:
  * Data visualization using ggplot
  * Numerical data analysis (measures of central tendency, measures of precision & uncertainty, quantiles)
  * Categorical data analysis (proportions)
* Perform a simple linear regression analysis and interpret the results in context of the problem



The `tidyverse` package is needed for these examples

```{r warning=FALSE, message=FALSE}
library(tidyverse)
```

## Palmer Penguins

The Palmer penguins dataset is a collection of body measurements for three penguin species that live on islands in the Palmer Archipelago in Antarctica: 

* Species: Adélie, Chinstrap, and Gentoo 
* Measurements: Bill length, bill depth, flipper length, body mass, island, sex, and year 
* Dataset creator: Dr. Kristen Gorman and the Palmer Station Long Term Ecological Research (LTER) team 
* Purpose: To study the relationship between Antarctic penguins' foraging behavior and environmental variability


### STEP 1: R Package

Install the R package and call it into your R environment.

```{r}
## INSTALL PACKAGE
#install.packages("palmerpenguins")

## LOAD THE LIBRARY
library(palmerpenguins)

## DATA
data("penguins")
```

### STEP 2: Look at the Data

Check out the structure of the data and take a look at it.

```{r}
## STRUCTURE 
str(penguins)

## HEAD, TAIL, OR VIEW
head(penguins)
```

Questions of interest: 

* How many penguins are in these data?
* What variables are included?  Are they numeric or categorical?

```{}
SPACE FOR ANSWERS:
```


## PART A: EDA for Categorical Variables

### STEP 3: One-way Tables and Bar Graphs

Make a table to show how many penguins there are of each species. 

```{r}
## ONE WAY TABLE FOR SPECIES
table(penguins$species)
```

What are your observations?

```{}
SPACE FOR ANSWERS:
```

Now make a bar Graph using ggplot2.

```{r}
## CUSTOM COLOR PALETTE
penCol<-c("darkorange","purple","cyan4")

## BAR GRAPH
ggplot(data = penguins, aes(x=species, fill=species))+
  geom_bar()+
  scale_fill_manual(values=penCol)
```

### STEP 4: Two-way Tables, Distributions, and Bar Graphs

Make a two-way table to show how many penguins there are of each species on each island. 

```{r}
## TWO WAY TABLE FOR SPECIES and ISLAND
tab2<-table(penguins$species, penguins$island)
tab2
```

Create a stacked bar graph to to show the total number of penguins from each island represented in the data set.  Fill by species. 

```{r}
## STACKED BAR GRAPH
ggplot(data = penguins, aes(x=island, fill=species))+
  geom_bar()+
  scale_fill_manual(values=penCol)
```


Now use the table to make conditional distributions for 

* 1) Distribution of location, given species 
* 2) Distribution of species, given location 

```{r}
## DISTRIBUTION OF LOCATION GIVEN SPECIES
prop.table(tab2, 1)

## DISTRIBUTION OF SPECIES GIVEN LOCATION
prop.table(tab2, 2)
```

Visualize the conditional distribution for species, given location 

```{r}
## FILLED BAR GRAPH
ggplot(data = penguins, aes(x=island, fill=species))+
  geom_bar(position = "fill")+
  scale_fill_manual(values=penCol)
```


Observations: 

Can this possibly tell you anything about how well different species have adapted to different environments or cohabitating? 

```{}
SPACE FOR ANSWERS:
```

## PART B: EDA for Numeric Variables 

### STEP 5: Numeric Summaries

Calculate the means and standard deviations body mass and flipper length for each species.  

```{r}
## MEANS AND ST DEV
## BODY MASS AND FLIPPER LENGTH
## BY SPECIES
penguins%>%
  group_by(species)%>%
  summarise(n=n(),
            avg_bodyMass=mean(body_mass_g, na.rm=TRUE), 
            sd_bodyMass=sd(body_mass_g, na.rm=TRUE), 
            avg_flipperLen=mean(flipper_length_mm, na.rm=TRUE), 
            sd_flipperLen=sd(flipper_length_mm, na.rm=TRUE))

```


### STEP 6: Boxplots, Histograms, and Density Plots 

Create plots to compare flipper length by species.

```{r}
## BOXPLOT
ggplot(data=penguins, aes(x=species, y=flipper_length_mm, fill=species))+
  geom_boxplot()+
  scale_fill_manual(values=penCol)+
  theme_minimal()

## HISTOGRAM (OVERLAPPING)
ggplot(data=penguins, aes(x=flipper_length_mm, fill=species))+
  geom_histogram(alpha = 0.5, 
                 position = "identity")+
  scale_fill_manual(values=penCol)+
  theme_minimal()

## DENSITY PLOT
ggplot(data=penguins, aes(x=flipper_length_mm, fill=species))+
  geom_density(alpha=0.5)+
  scale_fill_manual(values=penCol)+
  theme_minimal()

```


## PART C: EDA for Relationships Between Two Numeric Variables

Definition: 

Adaptive radiation is a process in evolutionary biology where a single species or small group of species rapidly diversifies into many new species. This process is characterized by the development of ecological and morphological diversity.

Let's explore if the three species have different beak (bill) morphologies.  

### STEP 7: Scatter Plot

Make a scatter plot for the relationship between bill length (x-axis) vs bill depth (y-axis). 

```{r}
## SCATTERPLOT 
## X: BILL LENGTH
## Y: BILL DEPTH

ggplot(data = penguins, aes(x=bill_length_mm, y=bill_depth_mm))+
  geom_point()+
  xlab("Bill length (mm)")+
  ylab("Bill depth (mm)")
```

Observations: 

```{}
SPACE FOR ANSWERS:
```


### STEP 8: Exploring Sugroups

```{r}
## SCATTERPLOT WITH SUBGROUPS 
## X: BILL LENGTH
## Y: BILL DEPTH
## COLOR: Species

ggplot(data = penguins, aes(x=bill_length_mm, y=bill_depth_mm, color=species))+
  geom_point()+
  scale_color_manual(values=penCol)+
  xlab("Bill length (mm)")+
  ylab("Bill depth (mm)")
```
 
## PART D: Modeling the Data

### STEP 9: Adding Smooths
```{r}
## GEOM_SMOOTH
ggplot(data = penguins, aes(x=bill_length_mm, y=bill_depth_mm, color=species))+
  geom_point()+
  scale_color_manual(values=penCol)+
  xlab("Bill length (mm)")+
  ylab("Bill depth (mm)")+
  geom_smooth(method="lm", se=FALSE)
```

### STEP 10: Create a Linear Model with Interactions

```{r}
## LINEAR MOD WITH INTERACTIONS
mod<-lm(bill_depth_mm~bill_length_mm*species, data=penguins)
summary(mod)
```

