---
title: 'DATA 151 BONUS Topic: Maps in R'
output:
  html_document:
    collapsed: no
    toc: yes
    toc_depth: 5
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '5'
---

### Learning Objectives

In this lesson students will learn how to create

* Choropleths (colored map plots)


## Choropleths (Map Plots)

### Example 1: All Trails

#### Step 1: Load the Data

```{r}
library(tidyverse)

npark <- read_csv("https://raw.githubusercontent.com/kitadasmalley/DATA151/main/Data/AllTrails%20data%20-%20nationalpark.csv")

#str(npark)
```

#### Step 2: State Level Data

Group by state to create summaries for metrics within a state. 
```{r}
stateNP<-npark%>%
  group_by(state_name)%>%
  summarise(stateTrails=##INSERT CODE##, 
            avgPop=##INSERT CODE##, 
            avgElev=##INSERT CODE##)

stateNP
```

#### Step 3: `usmap` Package

```{r}
#install.packages("usmap")
library(usmap)

states <- ##INSERT CODE##
  
#head(states)
```

Let's investigate the data for Oregon. 

##### Points
```{r}
oregon<-states%>%
  filter(full=="Oregon")

##INSERT CODE##
```

These data allow us to play "connect the dots" to draw the shape of the state of Oregon. 

##### Connect the dots

Oh no, what happened? 
```{r}
##INSERT CODE##
```

We need to tell R what order to connect the dots.

* `geom_path()` connects the observations in the order in which they appear in the data. 

* `geom_line()` connects them in order of the variable on the x axis.

```{r}
##INSERT CODE##
```

##### Filling in the space

We can actually think of geographies as generalized polygons!

```{r}
##INSERT CODE##
```


#### Step 4: Join the Map and Data

When joining the data to the map we need to have the same variable name in both.  Let's create a new column named `state_name`.

```{r}
## JOIN THE MAP AND THE DATA
##INSERT CODE##
```


#### Step 5: Make a Map

Our first map with the standard color scheme. 

```{r}
## OUR FIRST MAP!

##INSERT CODE##
```



#### STEP 6: Changing Color Palette

Viridis is a colorblind friendly color palette that can be used to create accessible heatmaps. 

```{r message=FALSE, warning=FALSE}
#install.packages("viridis")
library(viridis)

stateNP_Map%>%
  ggplot(aes(x, y, group = group)) +
  geom_polygon(aes(fill = stateTrails),color="black")+
  theme_bw()+
  coord_equal()+
  ggtitle("California has the MOST trails, but...")+
  scale_fill_viridis(option="viridis", direction = 1)

stateNP_Map%>%
  ggplot(aes(x, y, group = group)) +
  geom_polygon(aes(fill = avgPop),color="black")+
  theme_bw()+
  coord_equal()+
  ggtitle("..Oregon trails are the MOST popular")+
  scale_fill_viridis(option="viridis", direction = 1)

```

#### Your turn! 

Create maps to show the distribution of...

* Average elevation by state
* Average trail length by state


### Example 2: Dutch Bros

Dutch Bros Coffee is a drive-through coffee chain headquartered in Grants Pass, Oregon, with company-owned and franchise locations throughout the United States.

#### STEP 1: Dutch Bros Data

These data represent a list of store locations. There are variables for city, state, and zipcode.

```{r}
dutch<- read.csv("https://raw.githubusercontent.com/kitadasmalley/DATA151/main/Data/dutch_bros_locations.csv", 
                 header=TRUE)

str(dutch)
```
#### STEP 2: Summarise!

We will also want to wrangle our data so that we count the number of locations in a given state.

```{r}
## HOW MANY DUTCH BROS IN EACH STATE?
dutchSt<-##INSERT CODE##
```

#### STEP 3: Joining Data to the Map

When joining the data to the map we need to have the same variable name in both. Let’s create a new column named state in our states data, that contains the state abbreviations, since the dutch data set uses state abbreviations.

```{r}
## JOIN THE MAP TO THE DATA
mapDutch<-##INSERT CODE##
```

#### STEP 4: Choropleth

Dutch Bros on the west coast. 

```{r}
#install.packages("viridis")
library(viridis)

##INSERT CODE##
```

#### STEP 5: Interactivity 

```{}
#install.packages("plotly")
library(plotly)

p<-mapDutch%>%
  ggplot(aes(x, y, group = group)) +
  geom_polygon(aes(fill = n),color="black")+
  theme_bw()+
  coord_fixed()+
  scale_fill_viridis(option="viridis", direction = 1)

ggplotly(p)
```


### Example 3: Starbucks Worldwide


#### STEP 1: Starbucks Store Data

```{r}
starbucks <- read_csv("https://raw.githubusercontent.com/kitadasmalley/DATA151/main/Data/starbucks_directory.csv")

#str(starbucks)
```

##### How many counties is Starbucks in?

Let’s look at the format of the column Country:

```{r}
head(unique(starbucks$Country))
length(unique(starbucks$Country))
```

#### STEP 2: Using `map_data` in `ggplot2`

##### A. Load the data

```{r}
# install.packages("maps")
library(maps)

# Load map data
world_map <-  map_data("world")
```

##### B. Basic Map

A standard map with no adjustments.

```{r}
world_map%>% 
  ggplot(aes(map_id = region)) +
  geom_map(map = world_map)+
  expand_limits(x = world_map$long, y = world_map$lat)
```


##### C. Projections

The Mercator projection retains shapes of counties but distorts ratios. 

```{r}
#install.packages("mapproj")
library(mapproj)

## MERCATOR
world_map%>% 
  ggplot(aes(map_id = region)) +
  geom_map(map = world_map)+
  expand_limits(x = world_map$long, y = world_map$lat)+
  coord_map("mercator", xlim=c(-180,180)) 
```

The Mollweide projection preserves ratios. 

```{r}
## MOLLWEIDE
world_map%>% 
  ggplot(aes(map_id = region)) +
  geom_map(map = world_map)+
  expand_limits(x = world_map$long, y = world_map$lat)+
  coord_map("mollweide", xlim=c(-180,180)) 
```

#### STEP 3: Joining the Data to the Map

Let’s look at the map data. Note that region is used to specify country, but it has the whole name written out. This might be a problem for us because the Starbucks data uses a two-letter code of each country.

```{r}
head(world_map)
```

##### Use the countrycode Package

```{r}
#install.packages("countrycode")
library(countrycode)

starStores<-starbucks%>%
  group_by(Country)%>%
  summarise(stores=n())%>%
  mutate(region=countrycode(Country, origin = 'iso2c', destination = 'country.name'))%>%
  arrange(desc(stores))

head(starStores)
```

##### Join

```{r}
worldCoffee<-world_map%>%
  left_join(starStores, by="region")
```

#### STEP 4: Choropleth

We can use the ggthemes package for a map theme. 

```{r}
#install.packages("ggthemes")
library(ggthemes)

worldCoffee%>% 
  ggplot(aes(map_id = region, fill=stores)) +
  geom_map(map = world_map) +
  expand_limits(x = world_map$long, y = world_map$lat) +
  coord_map("mollweide", xlim=c(-180,180))+
  scale_fill_viridis(option="viridis", direction = 1)+
  theme_map()
```

##### Oh no! Something went wrong? Can you see it?

It turns out that only country name in maps that doesn’t use the whole country name is the United States. In the map packages the name is USA. 

##### Back to STEP 3

We can use the lapply function to apply any function to our data.

```{r}
## USE LAPPLY TO RECODE
starStores<-starbucks%>%
  group_by(Country)%>%
  summarise(stores=n())%>%
  mutate(region=as.character(lapply(Country, function(x){
    if(x!="US"){
      out=countrycode(x, origin = 'iso2c', destination = 'country.name')
    }
    if(x=="US"){
      out="USA"
      }
    out})))

worldCoffee2<-world_map%>%
  left_join(starStores, by="region")
```

##### Step 4 ... Again

Let's try that again..


```{r}
worldCoffee2%>% 
  ggplot(aes(map_id = region, fill=stores)) +
  geom_map(map = world_map) +
  expand_limits(x = world_map$long, y = world_map$lat) +
  coord_map("mollweide", xlim=c(-180,180))+
  scale_fill_viridis(option="viridis", direction = 1)+
  theme_map()
```

#### STEP 5: Number by City

Number of locations within the same city:

```{r}
## CITIES
starCities<-starbucks%>%
  group_by(Country, City)%>%
  summarise(lat=mean(Latitude), 
            long=mean(Longitude), 
            cityStores=n())

#str(worldCoffee2)
```

#### STEP 6: Add Points

Add points layer to plots 

```{r}
## ADD POINTS
worldCoffee2%>% 
  ggplot() +
  geom_map(map = world_map, aes(map_id = region, fill=stores)) +
   geom_point(data=starCities, aes(x=long, y=lat, size=cityStores), alpha=.3)+
  expand_limits(x = world_map$long, y = world_map$lat) +
  coord_map("mollweide", xlim=c(-180,180))+
  theme_map()+
  scale_fill_viridis(option="viridis", direction = 1)+
  guides(color = guide_legend(order=1),
          size = guide_legend(order=2))+
  theme(legend.position="bottom", legend.box = "vertical")
```


### Example 4: Vaccines in Oregon

The following data are available from the Oregon Health Authority

https://geo.maps.arcgis.com/apps/webappviewer/index.html?id=ea1c78a745c845d899a0184f3581a2ff

#### STEP 1: Load the 2023 - 2024 data into the R

```{r}
## FA24 DATA 502
## COUNTY LEVEL DATA
county<-read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/refs/heads/main/ProjectData/countyImmRate.csv")
#str(county)
#View(county)
```

#### STEP 2: Load maps package and map data

Get data at the county level.  Pay attention to the variables. 

```{r}
## LOAD MAP DATA
or_counties <- map_data("county", "oregon") %>% 
  select(lon = long, lat, group, id = subregion)

head(or_counties)
```

#### STEP 3: Join the map data to the vaccine data

The the tolower function to make strings lowercase. 

```{r}
## TO LOWER
##INSERT CODE##

# JOIN DATA
##INSERT CODE##
```

#### STEP 4: Map of Counties 

Map of vaccine rates across Oregon counties.

```{r}
library(viridis)

##INSERT CODE##
  
```

