---
title: "FALL 2024 - DataViz Midterm #2 - Part #3: Recreate"
author: "INSERT NAME HERE"
date: "DATE HERE"
output: html_document
---

### STEP 1: Visit the Oregon School Immunization Coverage (ArcGIS Map) for Preschools and childcare 

<https://geo.maps.arcgis.com/apps/webappviewer/index.html?id=ea1c78a745c845d899a0184f3581a2ff>

### STEP 2: Load the 2023 - 2024 data into the R

```{r}
## FA24 DATA 502
## COUNTY LEVEL DATA
county<-read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/refs/heads/main/ProjectData/countyImmRate.csv")
#str(county)

## VACCINE DATA FOR PRESCHOOLS
ps <- read.csv("https://raw.githubusercontent.com/kitadasmalley/Teaching/refs/heads/main/ProjectData/Preschool%20_%20Child%20Care.csv")

#View(ps)

## NEEDS TO BE NUMERIC
ps$Per_Vac_AllVac<-as.numeric(ps$Per_Vac_AllVac)
#str(ps)
```

### STEP 3: Load maps package and map data

```{r}
library(tidyverse)

## MAPS
#install.packages("maps")
library(maps)

## LOAD MAP DATA
### INSERT CODE HERE ###
```

### STEP 4: Join the map data to the vaccine data

```{r}
## JOIN MAPS AND DATA
### INSERT CODE HERE ###
```

### STEP 5: Recreate the choropleth map

```{r}
## USE geom_polygon() or geom_sf()
### INSERT CODE HERE ###
```

### STEP 6: Add layer for school points

```{r}
## ADD LAYER OF POINTS 
### INSERT CODE HERE ###
```

### STEP 7: Color Palettes
```{r}
## COLOR
# install.packages("colorspace")
library(colorspace)
hcl_palettes(plot = TRUE)
```

### STEP 8: Apply scales, themes, and guides to polish your graphic

```{r}
## POLISH
### INSERT CODE HERE ###
```


### STEP 9: Extra Credit for Interactivity

```{r}
## INTERACTION
### INSERT CODE HERE ###
```


### STEP 10: Extra Extra Credit for Ordering for Overlap

```{r}
## ORDER
### INSERT CODE HERE ###
```

### STEP 11: Make an alternative

Create an alternative that improves some aspect(s) of the graphic using best practices (improving on perception and/or color theory)


```{r}
## ALTERNATIVE
### INSERT CODE HERE ###
```

