---
title: 'DataViz: Interactive Map Activity'
author: "YOUR NAME HERE"
output: html_document
---


### Step 1: Connect to the Census API

First you will need to get a API key: <https://api.census.gov/data/key_signup.html>

Once you have an API key load it into your R environment so that you can access the ACS data.

```{r}
#install.packages("tidycensus")
library(tidycensus)

# YOUR CODE SHOULD LOOK LIKE THIS
# census_api_key("INCLUDE YOUR API HERE")
```

```{r echo=FALSE}
census_api_key("fbf65ba295ce81cab0a2eccd52d62ca564bf896b")
```

### Step 2: Variables available

Many many variables are included in the ACS. The ACS has 1 and 5 year estimates. Use the following code to see what variables are available. 

TASK: 
Pick a year and look at what estimates are available from the ACS 5-year estimates. 

```{r}
# Set a year of interest
this.year = 2010

# This looks at the 5 year estimates
# You can also do "acs1"
vars <- load_variables(year = this.year,
                      dataset = "acs5",
                      cache = TRUE)

# HOW MANY?
dim(vars)
```

Explore several possible explantory variables from the American Community Survey (ACS) including:

* B02001_001: Total
* B03002_003: White alone (Not Hispanic or Latino)
* B03002_004 Black or African American alone (Not Hispanic or Latino)
* B03002_012: Hispanic or Latino
* B03002_005: Native American alone (Not Hispanic or Latino)
* B03002_006: Asian alone (Not Hispanic or Latino)
* B03002_007: Native Hawaiian or Pacific Islander alone (Not Hispanic or Latino)
* B03002_009: Multiple Races (Not Hispanic or Latino)
* B03002_008: Other (Not Hispanic or Latino)
* B25064_001 MEDIAN GROSS RENT
* B25071_001: Rent Burden (MEDIAN GROSS RENT AS A PERCENTAGE OF HOUSEHOLD INCOME)
* B19013_001: MEDIAN HOUSEHOLD INCOME IN PAST 12 MONTHS
* B01002_001: Median age
* B25115_016: Renter Occupied - family
* B25115_027: Renter Occupied - nonfamily

### Step 3: Pick an Variable and a State
```{r}
## EXAMPLE
# MEDIAN HOME VALUE
orMedv <- get_acs(geography = "tract", year=this.year,
               state = "OR", #county = "Marion",
               variables = "B25077_001E", 
               geometry=TRUE)

head(orMedv)
```




### Step 4: Create a Map in Leaflet

TASKS: 
- Pick a color palette (change it from the one provided)
- Change the pop-up text

```{r}
#install.packages("leaflet")
library(leaflet)
library(tidyverse)


## WHAT TYPE OF COLOR PALETTE?
## Continuous?
pal<-colorNumeric("Greens", domain=0:ceiling(max(orMedv$estimate, na.rm=TRUE)))
## Quantile?
qpal<-colorQuantile("viridis", domain=orMedv$estimate,
                       n=5,na.color="#FFFFFF")

## CHANGE THE POP-UP TEXT
## TRY ADDING A LINE
popup<-paste("Tract: ", as.character(substring(orMedv$GEOID, 6, 11)), "<br>",
             "Median Home Value: ", as.character(orMedv$estimate))

leaflet()%>%
  addProviderTiles("CartoDB.Positron")%>%
  addPolygons(data=orMedv,
              fillColor= ~qpal(orMedv$estimate),
              fillOpacity = .7,
              weight =.5,
              smoothFactor = 0.2,
              popup = popup)%>%
      addLegend("bottomright", pal=qpal, values=orMedv$estimate,
                opacity = .7,
                title="Percentiles")

```

